NOAA Storm Database Exploration
======================================
Reproducible Research Peer Assessment 2
----------------------------------------
### Synopsis
This analysis started with two questions:

1. What type of events are the most harmful with respect to population health?
2. What type of events have the greatest economic impact?

In both cases, "TORNADO" was the clear leader in population health and economic impact. 

To access or learn about this data go to [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf).

### Data Processing
The overall approach here was to filter the data down to the fields needed and then apply calculations across Event Types for health impact and economic damage. 

The first step was to read in the data:
```{r}
## Read the data into R and subset it to the needed columns only
## assumes the data is in the working directory wiht original file name
tempdata <- read.csv("repdata-data-StormData.csv.bz2")
data <- tempdata[,c("STATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","CROPDMG")]
rm(tempdata)
```

The next step was to create a dataset for health:
```{r}
## Create a dataframe for injuries and fatalities by Event Type and 
## order the results. Uses a for loop through each EVTYPE and populates
## a new data frame called "health"
health <- data.frame()
Event <- levels(data$EVTYPE)
for (i in Event) {
    injury <- sum(data$INJURIES[data$EVTYPE == i])
    fatality <- sum(data$FATALITIES[data$EVTYPE == i])
    total <- injury + fatality
    nrow <- c(injury, fatality, total)
    health <- rbind.data.frame(health, nrow)
}
names(health) <- c("Injuries", "Fatalities", "Total")
health <- cbind.data.frame(Event, health)
## Order health by Total impact
health <- health[order(-health$Total), ]
```

Followed up by a dataset for economic impact:
```{r}
## Create a dataframe for property damage and crop damage by Event Type and 
## order the results. Uses a for loop through each EVTYPE and populates
## a new data frame called "health"
econ <- data.frame()
Event <- levels(data$EVTYPE)
for (i in Event) {
  propimpact <- sum(data$PROPDMG[data$EVTYPE == i])
  cropimpact <- sum(data$CROPDMG[data$EVTYPE == i])
  totalimpact <- propimpact + cropimpact
  nrow <- c(propimpact, cropimpact, totalimpact)
  econ <- rbind.data.frame(econ, nrow)
}
names(econ) <- c("PropDmg", "CropDmg", "TotalDmg")
econ <- cbind.data.frame(Event, econ)
# Order econ by Total Damage
econ <- econ[order(-econ$TotalDmg), ]
```

This concludes the data processing requirements for the analysis.

### Results
After running the processing we can plot and display our results. 

First we will look at the health impact:
```{r}
## Plot the top 10 results for health impact as a graph and data
tophealth <- health[1:10,]
qplot(Event,Total,data=tophealth,geom="bar",stat="identity",
  main="Top Ten Event Types for Health Impacts",xlab="Event Types",
  ylab="Total (Fatalities + Injuries)")
tophealth
```
We can see from the table and the plot that "TORNADO" is the clear leader in health impacts.

Next we will look at the economic impact:
```{r}
## Plot the top 10 results for economic impact as a graph and data
topecon <- econ[1:10,]
qplot(Event,TotalDmg,data=topecon,geom="bar",stat="identity",
      main="Top Ten Event Types for Economic Impacts",xlab="Event Types",
      ylab="Total (Property Damage + Crop Damage)")
topecon
```
Again, "TORNADO" is the leading contributor to economic impact across all severe weather events.

### Conclusion
Additional analysis to narrow down the types of events for specific geographic regions could be helpful in making perparedness or response efforts more effective. These results represent a national view of the data without taking into account how some states do not share in the impact of tornadoes. 